Practical Optional Types for Clojure

### 2015-10-17 formal-model.tex
It fits.
first submission
review first half

After demonstrating the core features Typed Clojure
provides, we link them together in a formal model called
\lambdatc{}.
%
Building on occurrence typing,
we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of syntax, typing rules, operational semantics,
and subtyping.

@@

\paragraph{Expressions} Syntax is given in \figref{main:figure:termsyntax}. Expressions \e{} 
include variables \x{}, values \v{},
applications, abstractions, conditionals, and let expressions.
All binding forms introduce fresh variables.
Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
integers {\nat{}},
constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
close value environments \openv{} over functions, which map bindings to values.

\paragraph{Types} Types \s{} or \t{} 
include the top type \Top,
\emph{untagged} unions {\Unionsplice {\overrightarrow{\t{}}}}, 
singletons ${\Value \singletonmeta{}}$,
and class instances \class{}.
We abbreviate the classes
\Booleanlong{} to \Boolean{}, 
\Keywordlong{} to \Keyword{},
\NumberFull{}  to \Number{},
\StringFull{}  to \String{}, and 
\FileFull{}  to \File{}.
We also abbreviate the types
\EmptyUnion{}     to \Bot{}, 
{\ValueNil}       to \Nil{}, 
{\ValueTrue}      to \True, and
{\ValueFalse} to {\False}.
%
The difference between the types
\Value{\Keyword} and \Keyword{} is subtle.
The former is inhabited by the class literal \Keyword{} and the result of 
\appexp{\classconst{}}{\makekw{a}}---the latter by keywords like \makekw{a}.
%
Function types 
$
{\ArrowOne {\x{}} {\s{}}
             {\t{}}
             {\filterset {\prop{}} {\prop{}}}
             {\object{}}}
$
contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions \prop{}, object \object{}, and return type
\t{},
which may refer to the function argument \x{}.
%Latent means they are relevant when the function is applied rather than evaluated.
They are instantiated with the
actual object of the argument in applications. % before they are used in the proposition environment.

\paragraph{Objects}
%As we saw in \secref{sec:overview},
%occurrence typing is capable of reasoning
%about deeply nested expressions.
To reason about nested expressions,
each expression is associated with 
a symbolic representation
called an \emph{object}.
%with respect to the current lexical environment. 
For example,
  variable \makelocal{m} has object \makelocal{m};
  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
  $42$ has the \emph{empty} object \emptyobject{}.
%
\figref{main:figure:termsyntax} gives the syntax for objects \object{}---non-empty objects 
\path{\pathelem{}}{\x{}} combine of a root variable \x{} and a \emph{path} \pathelem{},
which consists of
a possibly-empty sequence of \emph{path elements} (\pesyntax{}) applied right-to-left from the root variable.
We use two path elements---{\classpe{}} and {\keype{k}}---representing the results
of calling \classconst{} and of looking up a keyword $k$ respectively.

\paragraph{Propositions with a logical system}
In standard type systems, \emph{type environments}
track variables.
\begin{mathpar}
\infer [LC-Let]
{ \judgementtwo {{\propenv{}}}
                {\e{1}} {\s{}}
  \\
  \judgementtwo {{\propenv{}},\x{} \mapsto {\s{}}}
                {\e{2}} {\t{}}
}
{ 
  \judgementtwo {\propenv{}} 
            {\letexp{\x{}}{\e{1}}{\e{2}}} {\t{}}
           }

@@

\infer [T0-Local]
{ \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
{ \judgementtwo {\propenv{}} 
            {\x{}} {\t{}}
           }
\end{mathpar}
In T0-Local, 
$
{ \inpropenv {\propenv{}} {\isprop {\t{}} {\path{\pathelem{}}{\x{}}}} }
$
appeals to the proof system to solve \t{}.
%says under logical assumptions {\propenv{}}, object {\path{\pathelem{}}{\x{}}} is of type \t{}.
%We later define the more general T-Local.

@@

\paragraph{Methods} Next we convert {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
to
{\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
%The T-Method rule checks unresolved methods.
%$$
%    {\TMethod}
%$$
We use T-Method to check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}, 
which checks unresolved methods.
We verify the target type \s{} = \File is non-nil before erasing \Nil
by converting to a Java type \classhint{1} = \File.
The specific overload is chosen from the class table based on \classhint{1}---there is at most one.
Then {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
is checked against a nilable conversion of \classhint{1}, \t{1} = \Union{\Nil}{\File},
which succeeds by the previous rule.
The nilable return type \t{} = \Union{\Nil}{\String} is given, and---finally---the 
entire expression rewrites to expression \ref{eq:resolved}.
%
%We allow arguments to constructors and methods to be nilable, but not method
%and field targets.

@@

As an example,
\isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
has the true and false propositions 
\isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
meaning that if this expression produces true, \x{} is a keyword, otherwise it is not.

@@

\begin{listing}
$\letexp{hi_0} {\createmultiexp {\ArrowOne{\x{}}{\Keyword}{\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}} {\abs{\x{}}{\Keyword}{\x{}}}}
  {\\\text{\quad}
    \letexp{hi_1} {\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
      {\\\text{\quad\quad}
        \letexp{hi_2} {\extendmultiexp {hi_1} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}
        {\\\text{\quad\quad\quad
          \appexp{hi_2}{\makekw{en}}}}}}
$
\end{listing}

%For convenience, examples in this section are flattened when they are really nested
%let bindings. We also elide trivial latent propositions and objects.
%The following is an abbreviation of the previous expression.
%\begin{lstlisting}
%${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
%${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
%${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
%$\appexp{hi}{\makekw{en}}$
%\end{lstlisting}

%\defmethodliteral{} returns a new extended multimethod
%without changing the original multimethod. 
%
%\begin{minted}{clojure}
%(let [hi (defmulti [Kw -> Str] identity)]
%  (let [hi (defmethod hi :en [_] "hello")]
%    (let [hi (defmethod hi :fr [_] "bonjour")]
%      (hi :en))) ;=> "hello"
%\end{minted}

%\paragraph{How to check}
We now show to check
{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}.
%
The expression {\createmultiexp {\s{}} {\e{}}} creates a multimethod with \emph{interface type} \s{}, and dispatch function \e{}
of type \sp{},
producing a value of type
{\MultiFntype {\s{}} {\sp{}}}. % with interface type {\s{}} and dispatch function type {\sp{}}.
The T-DefMulti typing rule simply checks the dispatch function, and
verifies both the interface and dispatch type's domain agree.
%$$
%    \TDefMulti{}
%$$
Our example checks with \t{} = \Keyword, interface type \s{} = {\ArrowTwo{\x{}}{\Keyword}{\String}},
dispatch function type \sp{} = {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}, and overall type
$
{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
$.

@@

\begin{figure}
  \footnotesize
  $$
\LUpdate{}
$$
\caption{Update rule}
\label{main:figure:updaterule}
\end{figure}

@@

\paragraph{How to check}
A mandatory lookup is checked by T-GetHMap.
$$
\abs{\makelocal{b}}{\Breakfast}{\getexp{\makelocal{b}}{\makekw{en}}}
$$
The result type is \String, and the return object is \path{\keype{\makekw{en}}}{\makelocal{b}}.
The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
is a symbolic representation for a keyword lookup of $k$ in \object{}.
The substitution for {\x{}} handles the case where \object{} is empty.
$$
\begin{array}{rcl}
{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
\end{array}
$$

@@

The occurrence typing proof system uses standard propositional logic,
except for where nested information is combined. This is
handled by L-Update (\figref{main:figure:updaterule}).

@@

To demonstrate,
\makekw{lunch} meals hit the first HMap case and
update to {\HMapp {\extendmandatoryset {\mandatory{}}{\Valkw{Meal}}{\sp{}}} {\emptyabsent{}}}
where \sp{} = {\update{\Valkw{lunch}}{\Valkw{combo}}{\emptypath{}}}
and \mandatory{} = \mandatoryset{\mandatoryentry{\Valkw{Meal}}{\Valkw{lunch}},{\mandatoryentry{\Valkw{desserts}}{\Number{}}}}.
\sp{} updates to \Bot via the penultimate \updateliteral{} case,
because \restrict{\Value{\makekw{lunch}}}{\Value{\makekw{combo}}} = \Bot
by the first \restrictliteral{} case.
The same happens to \makekw{dinner} meals,
leaving just the \makekw{combo} HMap.

>>>>

After demonstrating the core features of Typed Clojure, 
we link them together in a formal model called
\lambdatc{}.
%
Building on occurrence typing,
we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of syntax, typing rules, operational semantics,
and subtyping.

@@

\paragraph{Expressions} Syntax is given in \figref{main:figure:termsyntax}. Expressions \e{} 
include variables \x{}, values \v{},
applications, abstractions, conditionals, and let expressions.
All binding forms introduce fresh variables---a subtle but important point since our type environments
are not simply dictionaries.
Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
integers {\nat{}},
constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
close value environments \openv{}---which map bindings to values---over functions.

\paragraph{Types} Types \s{} or \t{} 
include the top type \Top,
\emph{untagged} unions {\Unionsplice {\overrightarrow{\t{}}}}, 
singletons ${\Value \singletonmeta{}}$,
and class instances \class{}.
We abbreviate the classes
\Booleanlong{} to \Boolean{}, 
\Keywordlong{} to \Keyword{},
\NumberFull{}  to \Number{},
\StringFull{}  to \String{}, and 
\FileFull{}  to \File{}.
We also abbreviate the types
\EmptyUnion{}     to \Bot{}, 
{\ValueNil}       to \Nil{}, 
{\ValueTrue}      to \True, and
{\ValueFalse} to {\False}.
%
The difference between the types
\Value{\class{}} and \class{} is subtle.
The former is inhabited by class literals like \Keyword{} and the result of 
\appexp{\classconst{}}{\makekw{a}}---the latter by \emph{instances} of classes,
like a keyword literal \makekw{a}, an instance of the type \Keyword{}.
%
Function types 
$
{\ArrowOne {\x{}} {\s{}}
             {\t{}}
             {\filterset {\prop{}} {\prop{}}}
             {\object{}}}
$
contain \emph{latent} (terminology from~\cite{Lucassen88polymorphiceffect}) propositions \prop{}, object \object{}, and return type
\t{},
which may refer to the function argument \x{}.
%Latent means they are relevant when the function is applied rather than evaluated.
They are instantiated with the
actual object of the argument in applications. % before they are used in the proposition environment.

\paragraph{Objects}
%As we saw in \secref{sec:overview},
%occurrence typing is capable of reasoning
%about deeply nested expressions.
Each expression is associated with 
a symbolic representation
called an \emph{object}.
%with respect to the current lexical environment. 
For example,
  variable \makelocal{m} has object \makelocal{m};
  $\appexpone{\ccclass{\appexp{\makekw{lunch}}{\makelocal{m}}}}$ has object ${\path{\classpe{}}{\path{\keype{\makekw{lunch}}}{\makelocal{m}}}}$; and
  $42$ has the \emph{empty} object \emptyobject{}.
%
\figref{main:figure:termsyntax} gives the syntax for objects \object{}---non-empty objects 
\path{\pathelem{}}{\x{}} combine of a root variable \x{} and a \emph{path} \pathelem{},
which consists of
a possibly-empty sequence of \emph{path elements} (\pesyntax{}) applied right-to-left from the root variable.
We use two path elements---{\classpe{}} and {\keype{k}}---representing the results
of calling \classconst{} and looking up a keyword $k$, respectively.

\paragraph{Propositions with a logical system}
In standard type systems, association lists often
track the types of variables, like in LC-Let and LC-Local.
\begin{mathpar}
\infer [LC-Let]
{ \judgementtwo {{\propenv{}}}
                {\e{1}} {\s{}}
  \\
  \judgementtwo {{\propenv{}},\x{} \mapsto {\s{}}}
                {\e{2}} {\t{}}
}
{ 
  \judgementtwo {\propenv{}} 
            {\letexp{\x{}}{\e{1}}{\e{2}}} {\t{}}
           }

@@

\infer [T0-Local]
{ \inpropenv {\propenv{}} {\isprop {\t{}} {\x{}}}}
{ \judgementtwo {\propenv{}} 
            {\x{}} {\t{}}
           }
\end{mathpar}
In T0-Local, 
$
{ \inpropenv {\propenv{}} {\isprop {\t{}}{\x{}}}}
$
appeals to the proof system to solve for \t{}.
%says under logical assumptions {\propenv{}}, object {\path{\pathelem{}}{\x{}}} is of type \t{}.
%We later define the more general T-Local.

@@

\paragraph{Methods} Next we convert {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
to the non-reflective expression
{\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
%The T-Method rule checks unresolved methods.
%$$
%    {\TMethod}
%$$
We use T-Method to check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}, 
which checks unresolved methods.
We verify the target type \s{} = \File is non-nil before erasing \Nil
by converting to a Java type \classhint{1} = \File.
The specific overload is chosen from the class table based on \classhint{1}---there is at most one.
Then {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
is checked against a nilable conversion of \classhint{1}, \t{1} = \Union{\Nil}{\File},
which succeeds by the previous rule.
The nilable return type \t{} = \Union{\Nil}{\String} is given, and---finally---the 
entire expression rewrites to expression \ref{eq:resolved}.
%
%We allow arguments to constructors and methods to be nilable, but not method
%and field targets.

@@

As an example, the expression
\isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
has the true and false propositions 
\isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
meaning that if this expression produces true, \x{} is a keyword, otherwise it is not.

@@

\begin{listing}
$\letexp{hi_0} {\createmultiexp {\ArrowOne{\x{}}{\Keyword}{\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}} {\abs{\x{}}{\Keyword}{\x{}}}}
  {\\\text{\quad}
    \letexp{hi_1} {\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
      {\\\text{\quad\quad}
        \letexp{hi_2} {\extendmultiexp {hi_1} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}
        {\\\text{\quad\quad\quad
          \appexp{hi_2}{\makekw{en}}}}}}
$
\end{listing}
%
%For convenience, examples in this section are flattened when they are really nested
%let bindings. We also elide trivial latent propositions and objects.
%The following is an abbreviation of the previous expression.
%\begin{lstlisting}
%${\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}}$
%${\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}$
%${\extendmultiexp {hi} {\makekw{fr}} {\abs {\x{}} {\Keyword} {\makestr{bonjour}}}}$
%$\appexp{hi}{\makekw{en}}$
%\end{lstlisting}
%
%\defmethodliteral{} returns a new extended multimethod
%without changing the original multimethod. 
%
%\begin{minted}{clojure}
%(let [hi (defmulti [Kw -> Str] identity)]
%  (let [hi (defmethod hi :en [_] "hello")]
%    (let [hi (defmethod hi :fr [_] "bonjour")]
%      (hi :en))) ;=> "hello"
%\end{minted}
%
%\paragraph{How to check}
To check 
{\createmultiexp {\ArrowTwo{\x{}}{\Keyword}{\String}} {\abs{\x{}}{\Keyword}{\x{}}}},
%
we note
{\createmultiexp {\s{}} {\e{}}} creates a multimethod with \emph{interface type} \s{}, and dispatch function \e{}
of type \sp{},
producing a value of type
{\MultiFntype {\s{}} {\sp{}}}. % with interface type {\s{}} and dispatch function type {\sp{}}.
The T-DefMulti typing rule checks the dispatch function, and
verifies both the interface and dispatch type's domain agree.
%$$
%    \TDefMulti{}
%$$
Our example checks with \t{} = \Keyword, interface type \s{} = {\ArrowTwo{\x{}}{\Keyword}{\String}},
dispatch function type \sp{} = {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}, and overall type
$
{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
$.

@@

@@

\paragraph{How to check}
A mandatory lookup is checked by T-GetHMap.
$$
\abs{\makelocal{b}}{\Breakfast}{\getexp{\makelocal{b}}{\makekw{en}}}
$$
The result type is \String, and the return object is \path{\keype{\makekw{en}}}{\makelocal{b}}.
The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
is a symbolic representation for a keyword lookup of $k$ in \object{}.
The substitution for {\x{}} handles the case where \object{} is empty.
\begin{mathpar}
\begin{array}{rcl}
{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
\end{array}
\ \ \ \ \ \ \ 
\begin{array}{rcl}
{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
\end{array}
\end{mathpar}

@@

The occurrence typing proof system uses standard propositional logic,
except for where nested information is combined. This is
handled by L-Update:
{  \footnotesize
  $$
\LUpdate{}
$$
}

@@

To demonstrate,
\makekw{lunch} meals are handled by the first HMap case and
update to {\HMapp {\extendmandatoryset {\mandatory{}}{\Valkw{Meal}}{\sp{}}} {\emptyabsent{}}}
where \sp{} = {\update{\Valkw{lunch}}{\Valkw{combo}}{\emptypath{}}}
and \mandatory{} = \mandatoryset{\mandatoryentry{\Valkw{Meal}}{\Valkw{lunch}},{\mandatoryentry{\Valkw{desserts}}{\Number{}}}}.
\sp{} updates to \Bot via the penultimate \updateliteral{} case,
because \restrict{\Value{\makekw{lunch}}}{\Value{\makekw{combo}}} = \Bot
by the first \restrictliteral{} case.
The same happens to \makekw{dinner} meals,
leaving just the \makekw{combo} HMap.

### 2015-10-17 intro.tex
review first half

The popularity of dynamically-typed languages in software
development, combined with a recognition that types often improve
programmer productivity, software reliability, and performance, has
led to the recent development of a wide variety of optional and
gradual type systems aimed at checking existing programs written in
existing languages.  These include  TypeScript and Flow for
JavaScript, Hack for PHP, and MyPy
for Python among the optional systems, and Typed Racket, Reticulated
Python, and GradualTalk among gradually-typed systems.\footnote{We
  use ``gradual typing'' for systems like Typed Racket with sound
  interoperation between typed and untyped code; Typed Clojure or
 TypeScript which don't
  enforce type invariants we describe as ``optionally typed''.}
%FIXME: add citations for all those systems.

@@

We synthesize these lessons to present \emph{Typed Clojure}, an
optional type system for Clojure. 
%
Clojure is a dynamically
typed language in the Lisp family---built on the Java Virtual
Machine (JVM)---which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via
immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for parallelism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, it has an extensive base of existing untyped
programs, whose developers can benefit from Typed Clojure. As
a result, Typed Clojure is used in industry, experience we discuss
in this paper.

@@

But multimethods are not the only Clojure feature crucial to type
checking existing programs. As a language built on the Java Virtual
Machine, Clojure provides flexible and transparent access to existing
Java libraries, and \textbf{Clojure/Java interoperation} is found in almost
every significant Clojure code base. Typed Clojure therefore builds in
an understanding of the Java type system and handles interoperation
appropriately. Notably, \texttt{null} is a distinct type in Typed Clojure,
leading to automatic type-enforced
absence of null-pointer exceptions.

An example of these features is given in
figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
on the \texttt{class} of the argument---for \texttt{String}s,
the first method implementation is called, for \texttt{File}s, the
second. The \texttt{String} method calls
a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
\texttt{.getName} method 
on \texttt{File} requires a non-nil target, returning a nilable
type.  Typed Clojure offers an opt-in mode that
resolves JVM overloading, avoiding expensive runtime reflective calls.

Finally, high-performance immutable dictionaries
are the most common Clojure data structure,
which programmers repurpose liberally.
Simply treating them as uniformly-typed
key-value mappings would be insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous map} type, in which specific entries can be specified.

@@

\noindent
 The remainder of this paper begins with an example-driven
 presentation of the main type system features in
 \secref{sec:overview}. We then incrementally present a core calculus
 for Typed Clojure covering all of these features together in
 \secref{sec:formal} and prove type soundness
 (\secref{sec:metatheory}). We then 
 %discuss the full implementation of
 %Typed Clojure, \coretyped{}, which extends the formal model in many ways, 
 present an empirical analysis of significant code bases written
 in the full implementation of Typed Clojure---\coretyped{}---in \secref{sec:experience}. 
 Finally, we discuss related work and conclude.

>>>>

The popularity of dynamically-typed languages in software
development, combined with a recognition that types often improve
programmer productivity, software reliability, and performance, has
led to the recent development of a wide variety of optional and
gradual type systems aimed at checking existing programs written in
existing languages.  These include  TypeScript~\cite{typescript} and Flow~\cite{flow} for
JavaScript, Hack~\cite{hack} for PHP, and mypy~\cite{mypy}
for Python among the optional systems, and Typed Racket~\cite{TF08}, Reticulated
Python~\cite{Vitousek14}, and GradualTalk~\cite{gradualtalk} among gradually-typed systems.\footnote{We
  use ``gradual typing'' for systems like Typed Racket with sound
  interoperation between typed and untyped code; Typed Clojure or
 TypeScript which don't
  enforce type invariants we describe as ``optionally typed''.}
%FIXME: add citations for all those systems.

@@

We synthesize these lessons to present \emph{Typed Clojure}, an
optional type system for Clojure. 
%
Clojure is a dynamically
typed language in the Lisp family---built on the Java Virtual
Machine (JVM)---which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via
immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for parallelism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, there is an extensive base of existing untyped
programs whose developers can benefit from Typed Clojure,
an experience we discuss in this paper.

@@

But multimethods are not the only Clojure feature crucial to type
checking existing programs. As a language built on the Java Virtual
Machine, Clojure provides flexible and transparent access to existing
Java libraries, and \textbf{Clojure/Java interoperation} is found in almost
every significant Clojure code base. Typed Clojure therefore builds in
an understanding of the Java type system and handles interoperation
appropriately. Notably, \texttt{null} is a distinct type in Typed Clojure,
designed to automatically rule out null-pointer exceptions.

An example of these features is given in
\figref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
on the \texttt{class} of the argument---for \texttt{String}s,
the first method implementation is called, for \texttt{File}s, the
second. The \texttt{String} method calls
a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
\texttt{.getName} method 
on \texttt{File} requires a non-nil target, returning a nilable
type.  
%Typed Clojure offers an opt-in mode that
%resolves JVM overloading, avoiding expensive runtime reflective calls.

Finally, flexible, high-performance immutable dictionaries
are the most common Clojure data structure.
Simply treating them as uniformly-typed
key-value mappings would be insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous map} type, in which specific entries can be specified.

@@

\noindent
 The remainder of this paper begins with an example-driven
 presentation of the main type system features in
 \secref{sec:overview}. We then incrementally present a core calculus
 for Typed Clojure covering all of these features together in
 \secref{sec:formal} and prove type soundness
 (\secref{sec:metatheory}). We then 
 %discuss the full implementation of
 %Typed Clojure, \coretyped{}, which extends the formal model in many ways, 
 present an empirical analysis of significant code bases written
 in \coretyped{}---the full implementation of Typed Clojure---in \secref{sec:experience}. 
 Finally, we discuss related work and conclude.
